---
title: "Stat 585 - Data Types"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts, "tweaks.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: center, middle


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Types

---

## Basic Data Types

Name | Example
---- | -------
numeric | 2.93
integer | 2
character | "hello world"
logical | TRUE

Convert between data types with `as.XXX`, e.g. `as.numeric("2.394")` or `as.character("3.14159")`

Test whether something has a data type with `is.XXX`, e.g. `is.character("fuzzy puppies")` or `is.logical("abcde")`

---

## Basic Data Structures

Data structures hold one or more basic data types and may have multiple dimensions:

1D: Atomic vector (all one type) or List (multiple types)
2D: Matrix (all one type) or Data frame (multiple types)
nD: Array (all one type)

An individual value ("scalar" in other languages) is actualy a length one vector in R

---

## Basic Data Structures

`str()` can be used to examine a variable and determine its structure:

```{r}
library(tidyverse)
data("mtcars")
str(mtcars)

data("eurodist")
str(eurodist)

data("discoveries")
str(discoveries)

data("occupationalStatus")
str(occupationalStatus)

str(c(5:15))
str(letters)
str(3)
```
