---
title: "Stat 585 - Data Types"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts, "tweaks.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: center, middle


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Types

---

## Basic Data Types

Name | Example
---- | -------
numeric | 2.93
integer | 2
character | "hello world"
logical | TRUE

Convert between data types with `as.XXX`, e.g. `as.numeric("2.394")` or `as.character("3.14159")`

Test whether something has a data type with `is.XXX`, e.g. `is.character("fuzzy puppies")` or `is.logical("abcde")`

---

## Basic Data Structures

Data structures hold one or more basic data types and may have multiple dimensions:

   | One type | Multiple Types
-- | -------- | --------------
1D | Atomic vector | List
2D | Matrix | Data frame
nD | Array | 

An individual value ("scalar" in other languages) is a length one vector in R

---

## Basic Data Structures

`str()` can be used to examine a variable and determine its structure:

```{r, include = F}
library(tidyverse)
```

```{r, eval = -1}
library(tidyverse)
data("mtcars")
str(mtcars)
```
---

## Basic Data Structures

`str()` can be used to examine a variable and determine its structure:
```{r}
data("eurodist")
str(eurodist)
```
---

## Basic Data Structures

`str()` can be used to examine a variable and determine its structure:
```{r}
data("discoveries")
str(discoveries)
```
---

## Basic Data Structures

`str()` can be used to examine a variable and determine its structure:
```{r}
data("occupationalStatus")
str(occupationalStatus)
```
---

## Basic Data Structures

`str()` can be used to examine a variable and determine its structure:
```{r}
str(c(5:15))
str(month.name)
str(3)
str(mean)
```

---

## Factors

Some data has both names and a defined numerical ordering:
```{r}
month.name
```

Factors are used to store both the numerical values and the associated labels:
```{r}
factor(month.name)
```

By default, R uses alphabetical order for factors...

---

## Factors
```{r}
factor(month.name)
```

By default, R uses alphabetical order for factors...

Set level names to override the default ordering 
```{r}
months <- factor(month.name, levels = month.name)
months
```


---

## Factors

You can access the labels of a factor using `levels()`
```{r}
levels(months)
```

You can access the order of the factor using `as.numeric()`:
```{r}
as.numeric(months)
```

---

## Type Conversions
```{r}
# Get a numeric vector
x <- sample(1:10, 5)
x
# Create factor from numeric vector
fac_x <- factor(x)
fac_x

# Convert factor to numeric
as.numeric(fac_x) # Huh?
```
`as.numeric()` converts factors to their level numbers, not to their labels

---

## Type Conversions

```{r}
as.character(fac_x)
as.numeric(as.character(fac_x))
```
`as.character()` extracts the labels; you can then use `as.numeric` to get the labels' numeric value.
