---
title: "Stat 585 - elements of the tidyverse"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs    
    css: [default, default-fonts, "tweaks.css"]
    nature:
      beforeInit: "macros.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## tidyverse

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```


`tidyverse` is an 'opinionated' package bundling several other R packages for data science tasks:

```{r}
tidyverse::tidyverse_packages()
```

- share common data representations and API, i.e. work well together

- see https://github.com/hadley/tidyverse for more information


---

## Graphical and numerical summaries


1. graphical summaries - visualizations with `ggplot2`:

    - [RStudio cheatsheet for ggplot2](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
    
    - online documentation: https://ggplot2.tidyverse.org/
    
    - Google group: https://groups.google.com/forum/#!forum/ggplot2
  
2. numerical summaries - elements of `dplyr` and `tidyr`


---

## The pipe operator `%>%`

`f(x) %>% g(y)` is equivalent to `g(f(x), y)`

i.e. the output of one function is used as input to the next function. This function can be the identity

Consequences:

- `x %>% f(y)` is the same as `f(x, y)`

- statements of the form `k(h(g(f(x, y), z), u), v, w)` become
`x %>% 
   f(y) %>% 
   g(z) %>% 
   h(u) %>% 
   k(v, w)`

- read `%>%` as "then do"


---

## dplyr 

There are five primary `dplyr` *verbs*, representing distinct data analysis tasks:

- **Filter**: Select specified rows of a data frame, produce subsets

- **Arrange**: Reorder the rows of a data frame

- **Select**: Select particular columns of a data frame

- **Mutate**: Add new or change existing columns of the data frame (as functions of existing columns)

- **Summarise**: Create collapsed summaries of a data frame

- (**Group By**: Introduce structure to a data frame)

<br>

`dplyr` resources:

- https://dplyr.tidyverse.org/

- R for Data Science book: https://r4ds.had.co.nz/transform.html

---
class: inverse
## Your Turn

The Iowa State Police Department publishes a Daily Crime Log at http://www.police.iastate.edu/content/daily-crime-log

Data for 2017 and most of 2018 are available as file [pd-isu.csv](../data/pd-isu.csv) from the course website.

Read the data into your R session and answer the following questions:

- which types of crimes (`Classification`) are typically committed? How many different classifications are there?

- which days are in the top ten for 2018? (Use the `lubridate` package to get to date variables)

- how many times a day are crimes typically reported? 

- what else do you find?

---
class: middle, center
## 

Don't peek unless you have tried for yourself!

---

# a solution 

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
pd <- read_csv("../data/isu-pd.csv")

pd %>% count(Classification) %>% dim()
# 147 different types of crime classifications

pd  %>% count(Classification, sort = TRUE) %>% head()

```

---

# break multiple citings into Main and rest

```{r, warning= FALSE, message = FALSE}
pd <- pd %>% separate(Classification, into = "Main", sep="/", remove = FALSE) %>%
  mutate(
    Main = trimws(Main)
  )

pd  %>% count(Main, sort = TRUE) %>% head()
```

doesn't change much

---

## Top ten bad days in 2018

first get dates, see [RStudio lubridate cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/lubridate.pdf):
```{r}
head(pd$`Date/Time Reported`)

pd$date <- lubridate::mdy_hm(pd$`Date/Time Reported`)
summary(pd$date)
```

---

## Top ten bad days in 2018 (2)

Now we want to just have the day rather than the time

```{r}
pd$day <- lubridate::as_date(pd$date)
head(pd$day)
```


---

## Top ten bad days in 2018 (3)

```{r}
pd %>% 
  count(day, sort = TRUE) %>% 
  filter(lubridate::year(day) == 2018) %>%
  head(10)
```


---

## Average number of reports per day

Why is this average not right for the average number of reports by day?
```{r}
pd %>% count(day) %>% summary()
```

--

Problem: we are missing days with zero reports.

---
class: inverse
## Your Turn 


We want to find an elegant solution to the zero reports a day problem.

The function `complete` in the `tidyr` package looks promising.

Try to get it to work for this problem.

---
class: middle, center
##

... only after you tried ...

---

## Average number of reports per day (2)


Why is this still wrong?

```{r}
pd %>% 
  complete(day = full_seq(pd$day, period = 1)) %>% 
  count(day) %>%
  summary()
```

---

## Average number of reports per day (2)


What does `complete` do exactly?
```{r}
pd %>% 
  complete(day = full_seq(pd$day, period = 1))
```

---

## Average number of reports per day (3)

```{r}
perday <- pd %>% 
  complete(day = full_seq(pd$day, period = 1)) %>% 
  group_by(day) %>%
  summarize(
    n = sum(!is.na(Classification))
  ) 

perday %>%
  summary()
```

---

## Crime reports over time

```{r, fig.height = 5, fig.width = 8}
perday %>%
  mutate(year = lubridate::year(day)) %>%
  ggplot(aes( x = lubridate::yday(day), y = n)) + 
  geom_point() + 
  facet_grid(year~.)
```

---
class: inverse
## Your turn

The files [isu-football-2017.csv](../data/isu-football-2017.csv) and [isu-football-2018.csv](../data/isu-football-2018.csv) consist of the Cyclones' football schedule in 2017 and 2018. 

Use the data to create the chart below:

```{r, echo = FALSE, fig.height = 5, fig.width = 8}

schedule <- read.csv("../data/isu-football-2018.csv")
schedule <- schedule %>%
  mutate(
    date = lubridate::ymd("2018-01-01")
    )
lubridate::month(schedule$date) <- as.numeric(factor(schedule$Month, levels=c("Sep", "Oct", "Nov", "Dec"))) + 8
lubridate::mday(schedule$date) <- schedule$Day
schedule$yday <- lubridate::yday(schedule$date)
schedule$year <- lubridate::year(schedule$date)


schedule17 <- read.csv("../data/isu-football-2017.csv")
schedule17 <- schedule17 %>%
  mutate(
    date = lubridate::mdy(Date)
  )
schedule17$yday <- lubridate::yday(schedule17$date)
schedule17$year <- lubridate::year(schedule17$date)


perday %>%
  mutate(
    year = lubridate::year(day),
    yday = lubridate::yday(day)
    ) %>%
  filter(year <= 2018) %>%
  ggplot(aes(x = yday, y = n)) +
  geom_vline(data = schedule, aes(xintercept = yday, colour = home)) +
  geom_vline(data = schedule17, aes(xintercept = yday, colour = Home)) +
  geom_point(size = .75) +
  facet_grid(year~.) +
  theme_bw() +
  ylab("Number of police reports") +
  xlab("Day of the year")
```

Don't forget to *look at* the chart. What are your main findings?

---

## Tools for working with data

- moving between wide and long forms of data:

    - `gather`
    
    - `spread` 
    
    - see https://r4ds.had.co.nz/tidy-data.html


- joining data sets: `left_join`, `anti_join`, see also https://r4ds.had.co.nz/relational-data.html


---
class: inverse
## Your turn

Does the weather affect the number of citations? Use  [Ames_weather_2017-2018.csv](../data/Ames_weather_2017-2018.csv) and graphically examine the relationship between temperature and number of citations.

.small[For the curious: [script to get the weather data](https://gist.github.com/srvanderplas/9f4a471e0fb19495935fb2104f3ba9f8)]

```{r include = F}
library(lubridate)
ames_weather <- read_csv("../data/Ames_weather_2017-2018.csv") %>%
  filter(!is.na(time))

ames_weather_daily <- ames_weather %>%
  mutate(day = as.Date(date)) %>%
  group_by(day) %>%
  summarize(temp_low = min(temp_low), temp_high = max(temp_high))

# Just checking for weirdness
ggplot(data = ames_weather_daily) + 
  geom_ribbon(aes(x = day, ymin = temp_low, ymax = temp_high))


pd %>%
  filter(str_detect(Main, "Alcohol|Drug|Intox")) %>%
  mutate(Main = str_replace_all(Main, c("P[uU]blic Intox.*" = "PublicIntox",
                                        "Drug Violation" = "Drug",
                                        "Alcohol Violation" = "Alcohol",
                                        "Operating [wW]hile Intoxicated" = "OWI"))) %>%
  mutate(day = floor_date(date, unit = "day") %>% as.Date()) %>%
  group_by(Main, day) %>%
  summarize(n = n()) %>%
  spread(key = Main, value = n, fill = 0) %>%
  complete(day = full_seq(.$day, period = 1), fill = list(PublicIntox = 0, Drug = 0, Alcohol = 0, OWI = 0)) %>%
  left_join(ames_weather_daily) %>% 
  gather(key = Main, value = n, Alcohol:PublicIntox) %>%
  mutate(wday = wday(day, label = T)) %>%
  filter(wday %in% c("Thurs", "Fri", "Sat")) %>%
  ggplot(aes(x = day, y = n, size = temp_low)) + 
  geom_point(shape = 1) + 
  geom_smooth(aes(x = day, y = n), se = F, span = .1) + 
  facet_wrap(~Main, scales = "free_y") + 
  scale_size_continuous(range = c(0.05, 2))

```